---
title: '##Exploring the NOAA Storm Database: study on severe weather events'
author: "Wenjun Ma"
date: "12/4/2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

### Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

### Data
The data for this study come in the form of a comma-separeted-value file compressed via the bzip2 algorithm to reduce its size. The file is available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

### Synopsis

In this report, the following questions will be addressed:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

### Loading and Processing the data

##### Loading data into R
```{r}
data <- read.table("repdata-data-StormData.csv.bz2",header = TRUE, sep = ",",stringsAsFactors = FALSE)
head(data)
```
##### Extracting related data
```{r}
library("dplyr")
weather <- select(data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
colnames(weather) <- c("event.type","fatalities","injuries","property.damage","property.damage.exponent","crop.damage","crop.damage.exponent")
head(weather)
```
##### Impact on population health
```{r}
# Group and summarise data by fatalities and injuries
event.fatal.injur <- group_by(weather,event.type) %>% summarise(fatalities = sum(fatalities,na.rm=TRUE),injuries=sum(injuries,na.rm=TRUE))
# Rank weather events by fatalities and injuries from high to low
event.fatal <- arrange(event.fatal.injur,desc(fatalities))
event.injur <- arrange(event.fatal.injur,desc(injuries))
# Pick the top 10 harmful events
fatal10 <- event.fatal[1:10,]
injur10 <- event.injur[1:10,]
```
##### Impact on economic properties
```{r}
# Check values in property.damage.exponent
unique(weather$property.damage.exponent)
# Check values in crop.damage.exponent
unique(weather$crop.damage.exponent)
```
Please click [HERE](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html) for a brief explanation of "property.damage.exponent" and "crop.damage.exponent".
```{r}
# Create a function to convert values in property.damage.exponent and cop.damage.exponent to numeric values
exponent <- function(x){
      if (x %in% c("H","h"))
            return(10^2)
      else if (x %in% c("K","k"))
            return(10^3)
      else if (x %in% c("M","m"))
            return(10^6)
      else if (x %in% c("B","b"))
            return(10^9)
      else if (x == "+")
            return(10^0)
      else if (x %in% c("","-","?"))
            return(0)
      else
            return(10)
}
# Create columns to record full information about property and crop damage
property.expo <- sapply(weather$property.damage.exponent,FUN=exponent)
weather$damaged.property <- weather$property.damage * property.expo
crop.expo <- sapply(weather$crop.damage.exponent,FUN=exponent)
weather$damaged.crop <- weather$crop.damage * crop.expo
```
```{r}
# Group and summarise property and crop damage by events
event.property.crop <- group_by(weather,event.type) %>% summarise(property.damage = sum(damaged.property,na.rm=TRUE),crop.damage=sum(damaged.crop,na.rm=TRUE))
# Rank weather events by property and crop damage from high to low
event.property <- arrange(event.property.crop,desc(property.damage))
event.crop <- arrange(event.property.crop,desc(crop.damage))
# Pick the top 10 harmful events
property10 <- event.property[1:10,]
crop10 <- event.crop[1:10,]
```
### Results
1. Across the United States, which types of events are most harmful with respect to population health?
```{r}
print(fatal10[,1:2])
print(injur10[,c(1,3)])
par(mfrow=c(1,2),cex.main=0.8,cex.lab=0.8,cex.axis=0.8)
# Make a plot of the top 10 harmful events for causing fatalities
events.fatal <- fatal10$event.type
colors.fatal <- rainbow(10)
plot1 <- barplot(fatal10$fatalities,col=colors.fatal)
text(x = plot1, y = fatal10$fatalities, label = fatal10$fatalities, pos = 3, cex = 0.5)
text(plot1,par("usr")[1],labels=events.fatal,srt=40, adj = 1, xpd=TRUE,cex=0.7)
title(main="Top 10 harmful events for fatalities",ylab="Total number of fatalities")
# Make a plot of the top 10 harmful events for causing injuries
events.injur <- injur10$event.type
colors.injur <- rainbow(10)
plot2 <- barplot(injur10$injuries,col=colors.injur)
text(x = plot2, y = injur10$injuries, label = injur10$injuries, pos = 3, cex = 0.5)
text(plot2, par("usr")[1],labels=events.injur,srt=40, adj = 1, xpd=TRUE,cex=0.7)
title(main="Top 10 harmful events for injuries",ylab="Total number of Injuries")
```

The above plot shows that "tornado" was the most harmful weather event for both fatalities and injuries. Besides, it was much worse than any other severe weather events. It caused about 2 times more fatalities and about 12 times more injuries than the 2nd most harmful weather events for fatalities and injuries respectively.

2. Across the United States, which types of events have the greatest economic consequences?

```{r}
print(property10[,1:2])
print(crop10[,c(1,3)])
par(mfrow=c(1,2),cex.main=0.8,cex.lab=0.8,cex.axis=0.8)
# Make a plot of the top 10 harmful events for property damage
events.property <- property10$event.type
colors.property <- rainbow(10)
plot3 <- barplot(round(property10$property.damage/10^9,digits=2),col=colors.property)
text(x = plot3, y = round(property10$property.damage/10^9,digits=2), label = round(property10$property.damage/10^9,digits=2), pos = 3, cex = 0.5)
text(plot3,par("usr")[1],labels=events.property,srt=40, adj = 1, xpd=TRUE,cex=0.7)
title(main="Top 10 harmful events for property damage",ylab="Property damage in billion USD")
# Make a plot of the top 10 harmful events for causing injuries
events.crop <- crop10$event.type
colors.crop <- rainbow(10)
plot4 <- barplot(round(crop10$crop.damage/10^9,digits=2),col=colors.crop)
text(x = plot4, y = round(crop10$crop.damage/10^9,digits=2), label = round(crop10$crop.damage/10^9,digits=2), pos = 3, cex = 0.5)
text(plot4, par("usr")[1],labels=events.crop,srt=40, adj = 1, xpd=TRUE,cex=0.7)
title(main="Top 10 harmful events for crop damage",ylab="Crop damage in billion USD")
```

The above plot shows that "flood" caused most property damage, which was about twice of the second harmful weather event. "Drought" caused more crop damage than any other severe weather events.
